Investigating the molecular processes behind the cell-specific toxicity response to titanium dioxide nanobelts

### 2021-08-26 manuscript.tex
Push before applying all changes (manuscript accepted; we're going to use this for the proof changes).

@@

% Keywords
\keyword{nanomaterials; titanium dioxide; nanobelts; overrepresentation analysis; gene ontology; THP1; SAE; Caco2}

@@

In this study we re-examined a publicly available dataset generated by Tilton~\textit{et al.}~\cite{Tilton2013} (accession number: GSE42069)~\cite{Edgar2002} to study the detailed molecular mechanisms of TiO\textsubscript{2}-nanobelts toxicity. We analyzed gene expression data from three different cell lines exposed to either one of two concentrations of TiO\textsubscript{2}-nanobelts for 24 hours. The cell lines used were human primary small airway epithelial cells (SAE), human monocytic cells (THP1), and human epithelial colorectal adenocarcinoma cells (Caco2). These cell lines represent three common areas of exposure in the human body \textit{i.e.} skin epithelium, colon epithelium, and the innate immune system that typically will respond to particles that entered the body. We aim to provide an overview of the dose-dependent and cell type-specific response by focusing on toxicity-related processes. Apart from a significant increase in the activity of toxicity-related processes after exposure with a higher dosage, differences in intensity, and affected processes might occur between the cell lines. Using this work as an example, we will discuss the toxicological effects of TiO\textsubscript{2} nanobelts on the three cell lines and the importance and benefits of analysis workflows in the nanomaterial research field.

@@

\subsection*{Differential gene expression}
The differential gene expression analysis for the three cell lines after exposure to two different TiO\textsubscript{2}-nanobelt concentrations show that a high concentration of TiO\textsubscript{2}-nanobelts (Figure~\ref{fig:fig1}, right column) causes stronger gene expression changes in all three cell lines than at a low concentration (Figure~\ref{fig:fig1}, left column). While the Caco2 and SAE cells also show an increased response to the high TiO\textsubscript{2}-nanobelts concentration, the THP1 cells respond much stronger to the high concentration in terms of differentially expressed genes passing our criteria. Dose-dependent increases in response have been shown for many nanomaterials~\cite{Iavicoli2018,Ding2005}.

\begin{figure}[ht!]
\includegraphics[width=0.9\linewidth]{fig1.png}
  \caption{Gene expression volcano plots for different cell lines and TiO\textsubscript{2}-nanobelts concentrations.
   On the x-axis log\textsubscript{2}(fold change) is depicted whereas on the y-axis the -log\textsubscript{10}(p-value) is depicted. The dotted lines represent cut-off values for significantly changed genes (log\textsubscript{2} fold change $>$ 0.26 or $<$ -0.26, p-value $<$ 0.05). Brown color depicts that a gene meets both cut-off criteria, a blue color relates to meeting only the p-value cut-off, an orange color relates to meeting only the log2 fold change cut-off and grey color indicates that a gene does not meet any of the criteria.}
\label{fig:fig1}
\end{figure}

\subsection*{Pathway analysis}
Using the differentially expressed genes from the different cell lines, overrepresentation analysis was performed to identify altered pathways after TiO\textsubscript{2}-nanobelt exposure. The number of significantly overrepresented pathways (p-value $<$ 0.05) are shown in Table 1 in the column “Significant”. Concordant to the increase in differentially expressed genes matching our criteria shown in Figure 1, the number of resulting pathways increases with a surge in the concentration of TiO\textsubscript{2}-nanobelts. However, there was a smaller increase in differentially expressed genes, we found a similar increase in resulting pathways for the SAE cell line. While there was also an increase in the number of genes found for the Caco2 cell lines it did not directly translate to an increase in overrepresented pathways. Although over-representation analysis provides a great overview of all processes that are affected, it takes time to go manually over the many overrepresented pathways. Therefore, an automated method to select desired pathways, \textit{i.e.} toxicity-related pathways, gives a new approach to interpret the data.

\subsection*{Toxicity-related pathways}
To gain more insight into the toxicity-related processes, the overrepresented pathways were further categorized into pathways related to apoptotic process, inflammatory response, DNA damage response, and/or oxidative stress. Often this kind of clustering of the pathways is done manually. However, we implemented a gene-based approach. First, gene sets of four Gene~Ontology~(GO) terms were obtained, \textit{i.e.} “apoptotic process”~(GO:0006915, 1269 genes), “inflammatory response”~(GO:0006954, 569 genes),  “cellular response to DNA damage stimulus”~(GO:0006974, 762 genes), and “response to oxidative stress”~(GO:0006979, 243 genes). Evidently, these processes are tightly connected. Whereas some can be causative for others, they are expected to overlap.  Additional file A1 shows the gene overlap between the four GO gene sets in a Venn-diagram (see Additional file A1). Next, we calculated the gene overlap between the pathways from the WikiPathways pathway collection with the annotated genes of the four GO-terms. The overlap was calculated by dividing the number of toxicity-related genes present in the pathway by total number of genes present in the pathway. As a cut-off we used 50$\%$ indicating that at least half of the pathway is directly linked to GO-term of interest via their gene overlap.

To select the desired gene overlap cut-off, we compared four cut-offs with each other and an overrepresentation analysis approach to see how many pathways these would give. The cut-offs we used were 50\%, 60\%, 70\% and 80\%. For the overrepresentation analysis we looked for overrepresentation of the pathway genes in the gene lists of the GO-terms. The results are shown in Table~A2 (see Additional file A2). The overrepresentation analysis (p-value < $<$ 0.05) showed the most pathways to be linked to the four GO-terms, in this case it showed 401, 250, 203 and 205 pathways overrepresented for the apoptotic process, inflammatory response, DNA damage and oxidative stress GO-terms respectively. The 80\% and 70\% gene overlap resulted in almost similar amounts of pathways compared to each other. They even resulted in 0 pathways for the inflammatory response GO-term. Moreover, it stands out that for the gene overlap approach only one pathway meets at least 50\% or higher gene overlap with the oxidative stress GO-term. Additionally, it is interesting to see that an overrepresentation analysis prompts hundreds of pathways compared to just dozens when looking at gene overlap. Comparing the results of the overrepresentation analysis and various cut-offs the at least 50\% gene overlap cut-off was deemed best for our approach.

@@

The Caco2 cell line shows for both concentration 10 overrepresented pathways. While the SAE cell line shows 3 for the low concentration, which could be an indication of a very little toxic response to the TiO\textsubscript{2}-nanobelts exposure, and 15 for the high concentration. The THP1 cell line shows a clear toxic response activation for the high concentration with 39 overrepresented pathways, and 10 for the low concentration.

\begin{table}[]
\caption{Table depicting the number of significantly overrepresented pathways, altered toxicity pathways and the number of altered toxicity pathways for each GO-term.}
\resizebox{\columnwidth}{!}{%
\begin{tabular}{lcc|cccc}
 & \textbf{Significant} & \textbf{Toxicity} & \textbf{Apoptosis} & \textbf{Inflammation} & \textbf{\begin{tabular}[c]{@{}c@{}}DNA \\ damage\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}Oxidative \\ stress\end{tabular}} \\ \hline
\textbf{Pathways} & 1076 & 101 & 66 & 15 & 28 & 2 \\ \hline
Caco2 low & 56 & 10 & 5 & 0 & 5 & 0 \\
Caco2 high & 58 & 10 & 10 & 1 & 0 & 0 \\ \hline
SAE low & 28 & 3 & 1 & 0 & 2 & 0 \\
SAE high & 50 & 15 & 6 & 1 & 8 & 0 \\ \hline
THP1 low & 38 & 10 & 9 & 4 & 0 & 0 \\
THP1 high & 201 & 39 & 31 & 9 & 3 & 1 
\end{tabular}%
}
\end{table}

@@

\subsection*{Caco2 cells}
While for both concentrations the Caco2 cell line shows the same number of overrepresented pathways, the pathways compared between the low- and high concentration are not all the same. Of the ten pathways, three are found for both concentration, namely IL-5 Signaling Pathway (wikipathways:WP127,~\cite{WP127}), IL-2 Signaling Pathway (wikipathways:WP49,~\cite{WP49}) and Endometrial cancer (wikipathways:WP4155,~\cite{WP4155}). Interestingly to see is that for the low concentration pathways related to DNA-damage and repair show up in the results. For example, HDR through Homologous Recombination (HRR) or Single Strand Annealing (SSA) (wikipathways:WP3567,~\cite{WP3567}), Nonhomologous End-Joining (NHEJ) (wikipathways:WP3550,~\cite{WP3550}) and DNA Double Strand Break Response (wikipathways:WP3543,~\cite{WP3543}) are amongst these pathways. While for the high concentration it is interesting to see that the Apoptosis pathway (wikipathways:WP254,~\cite{WP254}) shows up. Which could be an indication of Caco2 cells possibly undergoing apoptotic processes upon exposure to TiO\textsubscript{2}-nanobelts. However, These results indicate that the studied processes in Caco2 cell lines are not extensively affected by the administration of TiO\textsubscript{2}-nanobelts, for either the low or the high concentration. This finding supports the conclusion based on the cell viability assay results in the original study which showed that there was no significant decrease in cell viability for the Caco2 cell line for both concentrations of TiO\textsubscript{2}-nanobelts~\cite{Tilton2013}.

@@

@@

Caco2 cells seem resilient to the exposure and show very little activation of toxicity-related processes. The passage of nanoparticles through the cellular barriers of Caco2 cells has been shown to be limited which could result in reduced uptake and therefore less toxic response~\cite{Ye2017}. Subsequently, this indicates a lower importance of gastro-intestinal uptake in general. On the other hand, TiO\textsubscript{2}-nanobelts caused the biggest toxicological response to THP1 cells,as they are more sensitive to exposure compared to epithelial cell lines RLE-6TN and BEAS-2B~\cite{Xia2013}. Moreover, it has been reported that this response is specific to the nanobelt form of TiO\textsubscript{2}~\cite{Xia2013}.

@@

To illustrate how the effects can be studied in more depth we visualized the log fold change of all six conditions on the Oxidative Stress pathway (wikipathways:WP408~\cite{WP408}), the Apoptosis pathway (wikipathways:WP254~\cite{WP254}) and the DNA Mismatch Repair pathway (wikipathways:WP531~\cite{WP531}). It shows that for most genes in these pathways gene expression data is present (see Additional file A3 and A4).

While the Oxidative Stress pathway shown in Figure~\ref{fig:fig3}, only shows up to be significantly overrepresented for the THP1 cell line for the high concentration it is still an interesting pathway to dive deeper into. The NFKB1 gene, which encodes for the Nf$\kappa$B-p105 subunit, has the highest log fold change for the THP1 cell line, high concentration. Additionally, the THP1 cell line, low concentration shows a positive log fold change as well, whereas the SAE cell line shows negative log fold changes and the Caco2 cell line noticeably lower positive log fold changes for both concentrations. Furthermore, the SOD2 gene, which is involved in the conversion of superoxide and protects against oxidative stress, shows a similar pattern~\cite{Urso2003}. These genes indicate that the THP1 cell line responds to oxidative stress by increasing the expression of protective genes.

\begin{figure}[ht!]
\includegraphics[height=12cm,keepaspectratio]{fig3.png}
  \caption{Visualization of log fold change on the Oxidative Stress pathway (wikipathways:WP408) for all six conditions.
   Cell lines are depicted from left to right as Caco2, SAE and THP1. The top row depicts the low concentration whereas the bottom row depicts the high concentration. Gradient goes from blue (log fold change $<$ -0.58) via white (log fold change = 0.0) to red (log fold change $>$ 0.58).
}
\label{fig:fig3}
\end{figure}

Moreover, the Apoptosis pathway shows positive log fold changes of the CASP2 and CASP7 genes, which are involved in apoptosis execution, for the SAE and THP1 cell lines for both concentrations whereas the Caco2 cell line shows no negative or positive log fold change~\cite{BouchierHayes2011,Lamkanfi2010}. Furthermore, the pathway shows positive log fold changes for the apoptosis promoting interferon regulatory factors such as IRF4 and IRF5~\cite{Stang2007,Fanzo2003,Hu2009}, which is a similar pattern as described for the CASP2 and CASP7 genes. This could be an indication that apoptosis is stimulated in these cell lines. However, the original study by Tilton \textit{et al.} does not show a significant decrease in cell viability. 
The DNA Mismatch Repair pathway shows a positive log fold change for the Caco2 THP1 cell lines for the LIG1 gene, which encodes for DNA ligase 1. This enzyme is involved in both DNA replication and in this context more importantly repair~\cite{Mota2019}. The aforementioned genes show positive log fold changes for the THP1 cell line throughout the pathway. The increase in expression of these genes could be an indication that DNA mismatch repair is activated due to exposure to TiO\textsubscript{2}-nanobelts.

\subsection*{The advantage of our approach}
Enrichment analysis for gene expression data is well-established and can be easily automated to increase reproducibility~\cite{Yu2012}. The interpretation of the long list of altered, often overlapping pathways is still a challenge. By providing an automated approach to filter the pathway enrichment result towards a specific biological focus, in this case toxicity, a more context-specific interpretation is facilitated. The generated pathway-gene networks showcase the connectedness of the processes and provide a more systemic view than looking at individual pathways. In addition, the approach yields a fast and easy method to select pathways of interest for more detailed scrutiny as we have shown. While interpretation and comparison between multiple datasets on a process level are still challenging, the automated analysis workflow used supports the exploration of the data~\cite{TiO2-scripts}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}

This study investigated the molecular response of three different cell lines to exposure to TiO\textsubscript{2}-nanobelts. Using our process-level analysis based on pathway analysis, pathway selection using gene sets, and network visualization, our findings support the results by S.C. Tilton \textit{et al.} showing that the three cell lines, Caco2, SAE, and THP1, show different toxicity-related responses to the exposure of TiO\textsubscript{2}-nanobelts from resilient Caco2 and SAE cells to strongly responding THP1 cells. The latter is not unexpected since the observed effects align with the biological function of these immune cells. Importantly, the approach allowed us to not just find changes in gene expression, but also responding molecular pathways via the pathway analysis and, additionally, allows us to filter the broad pathway enrichment results to a focused view on the cytotoxic processes affected. This allows us to visualize and explore the interactions between responding genes, based on underlying molecular processes. This versatile approach captured in the R-script can be easily adapted to isolate other processes by using other Gene Ontology terms diverting the focus to other biological processes of interest.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}

@@

The dataset consisted of 18 samples from three human-like cell lines, \textit{i.e.} Caco2, SAE, and THP1, which were exposed to either medium (control), 10~$\mu$g/ml or 100~$\mu$g/ml TiO\textsubscript{2}-nanobelts for 24 hours in triplicate. Culture conditions were kept as identical as possible between the three cell lines. Details can be found via the original publication by Tilton~\textit{et al.}~\cite{Tilton2013}.

@@

\subsection*{Toxicity-related processes}
Within the two human pathway collections used, toxicity-related pathways were identified based on the gene overlap with the toxicity-related gene sets. The overlap was calculated by dividing the number of toxicity-related genes present in the pathway by total number of genes present in the pathway. The gene sets were retrieved from the Gene Ontology (GO, version: release 2020-06) for the following four toxicity-related GO-terms: “apoptotic process” (GO:0006915), “inflammatory response” (GO:0006954), “cellular response to DNA damage stimulus” (GO:0006974) and “response to oxidative stress” (GO:0006979)~\cite{Ashburner2000,Carbon2020}. Associated genes were retrieved using the biomaRt package in R (version 2.42.0, Ensembl Genes 100)~\cite{Durinck2005,Durinck2009}. Using the GO evidence codes, only genes with experimental evidence or manually curated annotations were included to ensure high confidence that the gene is involved in the specific process (IBA, IC, IDA, IEP, IGI, IMP, IPI, TAS, http://geneontology.org/docs/guide-go-evidence-codes/, see Additional file A5). Gene overlap between GO-terms was visualized using Venny version 2.1.0~\cite{Venny}.

@@

\noindent 
\begin{tabular}{@{}ll}
TiO\textsubscript{2} & titanium dioxide\\
SAE & small airway epithelial cells\\
THP1 & human monocytic cells\\
Caco2 & human epithelial colorectal adenocarcinoma cells\\
GO & Gene Ontology\\
\end{tabular}}

>>>>

\usepackage{xcolor}
\usepackage{soul}

@@

% Keywords
\keyword{nanomaterials; titanium dioxide; nanobelts; overrepresentation analysis; gene ontology; THP1; SAE; Caco-2/HT29-MTX}

@@

In this study we re-examined a publicly available dataset generated by Tilton~\textit{et al.}~\cite{Tilton2013} (accession number: GSE42069)~\cite{Edgar2002} to study the detailed molecular mechanisms of TiO\textsubscript{2}-nanobelts toxicity. We analyzed gene expression data from three different \hl{cellular models} exposed to either one of two concentrations of TiO\textsubscript{2}-nanobelts for 24 hours. The cell lines used were human primary small airway epithelial cells (SAE), human monocytic cells (THP1), and human epithelial colorectal adenocarcinoma cells \hl{co-cultured with HT29-MTX goblet cells (Caco-2/HT29-MTX)}. These \hl{cellular models} represent three common areas of exposure in the human body \textit{i.e.} \hl{small airway} epithelium, colon epithelium, and the innate immune system that typically will respond to particles that entered the body. We aim to provide an overview of the dose-dependent and cell type-specific response by focusing on toxicity-related processes. Apart from a significant increase in the activity of toxicity-related processes after exposure with a higher dosage, differences in intensity, and affected processes might occur between the cell lines. Using this work as an example, we will discuss the toxicological effects of TiO\textsubscript{2} nanobelts on the three \hl{cellular models} and the importance and benefits of analysis workflows in the nanomaterial research field.

@@

\subsection*{Differential gene expression}
The differential gene expression analysis for the three cell lines after exposure to two different TiO\textsubscript{2}-nanobelt concentrations show that a high concentration of TiO\textsubscript{2}-nanobelts (Figure~\ref{fig:fig1}, right column) causes stronger gene expression changes in all three \hl{cellular models} than at a low concentration (Figure~\ref{fig:fig1}, left column). While the \hl{Caco-2/HT29-MTX} and SAE cells also show an increased response to the high TiO\textsubscript{2}-nanobelts concentration, the THP1 cells respond much stronger to the high concentration in terms of differentially expressed genes passing our criteria. Dose-dependent increases in response have been shown for many nanomaterials~\cite{Iavicoli2018,Ding2005}.

\begin{figure}[ht!]
\includegraphics[width=0.9\linewidth]{fig1.png}
  \caption{Gene expression volcano plots for different \hl{cellular models} and TiO\textsubscript{2}-nanobelts concentrations.
   On the x-axis log\textsubscript{2}(fold change) is depicted whereas on the y-axis the -log\textsubscript{10}(p-value) is depicted. The dotted lines represent cut-off values for significantly changed genes (log\textsubscript{2} fold change $>$ 0.26 or $<$ -0.26, p-value $<$ 0.05). Brown color depicts that a gene meets both cut-off criteria, a blue color relates to meeting only the p-value cut-off, an orange color relates to meeting only the log2 fold change cut-off and grey color indicates that a gene does not meet any of the criteria.}
\label{fig:fig1}
\end{figure}

\subsection*{Pathway analysis}
Using the differentially expressed genes from the different \hl{cellular models}, overrepresentation analysis was performed to identify altered pathways after TiO\textsubscript{2}-nanobelt exposure. The number of significantly overrepresented pathways (p-value $<$ 0.05) are shown in Table 1 in the column “Significant”. Concordant to the increase in differentially expressed genes matching our criteria shown in Figure 1, the number of resulting pathways increases with \hl{an increase} in the concentration of TiO\textsubscript{2}-nanobelts. \hl{While} there was a smaller increase in differentially expressed genes, we found a similar increase in resulting pathways for the SAE cell line. There was also an increase in the number of genes found for the \hl{Caco-2/HT29-MTX co-culture, however,} it did not directly translate to an increase in overrepresented pathways. Although overrepresentation analysis provides a great overview of all processes that are affected, it takes time to go manually over the many overrepresented pathways. Therefore, an automated method to select desired pathways, \textit{i.e.} toxicity-related pathways, gives a new approach to interpret the data.

\subsection*{Toxicity-related pathways}
To gain more insight into the toxicity-related processes, the overrepresented pathways were further categorized into pathways related to apoptotic process, inflammatory response, DNA damage response, and/or oxidative stress. Often this kind of clustering of the pathways is done manually. However, we implemented a gene-based approach. First, gene sets of four Gene~Ontology~(GO) terms were obtained, \textit{i.e.} “apoptotic process”~(GO:0006915, 1269 genes), “inflammatory response”~(GO:0006954, 569 genes),  “cellular response to DNA damage stimulus”~(GO:0006974, 762 genes), and “response to oxidative stress”~(GO:0006979, 243 genes). Evidently, these processes are tightly connected. Whereas some can be causative for others, they are expected to overlap. Additional Figure \ref{fig:figA1} shows the gene overlap between the four GO gene sets in a Venn-diagram. Next, we calculated the gene overlap between the pathways from the WikiPathways pathway collection with the annotated genes of the four GO-terms. The overlap was calculated by dividing the number of toxicity-related genes present in the pathway by total number of genes present in the pathway. As a cut-off we used 50$\%$ indicating that at least half of the pathway is directly linked to GO-term of interest via their gene overlap.

To select the desired gene overlap cut-off, we compared four cut-offs with each other and an overrepresentation analysis approach to see how many pathways these would give. The cut-offs we used were 50\%, 60\%, 70\% and 80\%. For the overrepresentation analysis we looked for overrepresentation of the pathway genes in the gene lists of the GO-terms. The results are shown in additional Table \ref{tab:tabA1}. The overrepresentation analysis (p-value < $<$ 0.05) showed the most pathways to be linked to the four GO-terms, in this case it showed 401, 250, 203 and 205 pathways overrepresented for the apoptotic process, inflammatory response, DNA damage and oxidative stress GO-terms respectively. The 80\% and 70\% gene overlap resulted in almost similar amounts of pathways compared to each other. They even resulted in 0 pathways for the inflammatory response GO-term. Moreover, it stands out that for the gene overlap approach only one pathway meets at least 50\% or higher gene overlap with the oxidative stress GO-term. Additionally, it is interesting to see that an overrepresentation analysis prompts hundreds of pathways compared to just dozens when looking at gene overlap. Comparing the results of the overrepresentation analysis and various cut-offs the at least 50\% gene overlap cut-off was deemed best for our approach.

@@

The \hl{Caco-2/HT29-MTX co-culture} shows for both concentration 10 overrepresented pathways. While the SAE \hl{cells} show 3 for the low concentration, which could be an indication of a very little toxic response to the TiO\textsubscript{2}-nanobelts exposure, and 15 for the high concentration. The THP1 cell line shows a clear toxic response activation for the high concentration with 39 overrepresented pathways, and 10 for the low concentration.

\begin{table}[]
\caption{Table depicting the number of significantly overrepresented pathways, altered toxicity pathways and the number of altered toxicity pathways for each GO-term.}
\resizebox{\columnwidth}{!}{%
\begin{tabular}{lcc|cccc}
 & \textbf{Significant} & \textbf{Toxicity} & \textbf{Apoptosis} & \textbf{Inflammation} & \textbf{\begin{tabular}[c]{@{}c@{}}DNA \\ damage\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}Oxidative \\ stress\end{tabular}} \\ \hline
\textbf{Pathways} & 1076 & 101 & 66 & 15 & 28 & 2 \\ \hline
\hl{Caco-2/HT29-MTX} low & 56 & 10 & 5 & 0 & 5 & 0 \\
\hl{Caco-2/HT29-MTX} high & 58 & 10 & 10 & 1 & 0 & 0 \\ \hline
SAE low & 28 & 3 & 1 & 0 & 2 & 0 \\
SAE high & 50 & 15 & 6 & 1 & 8 & 0 \\ \hline
THP1 low & 38 & 10 & 9 & 4 & 0 & 0 \\
THP1 high & 201 & 39 & 31 & 9 & 3 & 1 
\end{tabular}%
}
\end{table}

@@

\subsection*{\hl{Caco-2/HT29-MTX} cells}
While for both concentrations the \hl{Caco-2/HT29-MTX co-culture} shows the same number of overrepresented pathways, the pathways compared between the low- and high concentration are not all the same. Of the ten pathways, three are found for both concentration, namely IL-5 Signaling Pathway (wikipathways:WP127,~\cite{WP127}), IL-2 Signaling Pathway (wikipathways:WP49,~\cite{WP49}) and Endometrial cancer (wikipathways:WP4155,~\cite{WP4155}). Interestingly to see is that for the low concentration pathways related to DNA-damage and repair show up in the results. For example, HDR through Homologous Recombination (HRR) or Single Strand Annealing (SSA) (wikipathways:WP3567,~\cite{WP3567}), Nonhomologous End-Joining (NHEJ) (wikipathways:WP3550,~\cite{WP3550}) and DNA Double Strand Break Response (wikipathways:WP3543,~\cite{WP3543}) are amongst these pathways. While for the high concentration it is interesting to see that the Apoptosis pathway (wikipathways:WP254,~\cite{WP254}) shows up. Which could be an indication of \hl{Caco-2/HT29-MTX} cells possibly undergoing apoptotic processes upon exposure to TiO\textsubscript{2}-nanobelts. However, These results indicate that the studied processes in \hl{Caco-2/HT29-MTX co-cultures} are not extensively affected by the administration of TiO\textsubscript{2}-nanobelts, for either the low or the high concentration. This finding supports the conclusion based on the cell viability assay results in the original study which showed that there was no significant decrease in cell viability for the \hl{Caco-2/HT29-MTX co-culture} for both concentrations of TiO\textsubscript{2}-nanobelts~\cite{Tilton2013}.

@@

\hl{The THP1 cell line is the cellular model which has the most inflammation related pathways in the results, four for the low concentration and nine for the high concentration. All four of the low concentration show up in the results of the high concentration as well. These pathways are: Photodynamic therapy-induced NF-kB survival signaling} (wikipathways:WP3617,~\cite{WP3617}), \hl{Interleukin-10 signaling} (wikipathways:WP4063,~\cite{WP4063}), \hl{Fibrin Complement Receptor 3 Signaling Pathway} (wikipathways:WP4136,~\cite{WP4136}) \hl{and Platelet-mediated interactions with vascular and circulating cells} (wikipathways:WP4462,~\cite{WP4462}).
\hl{Nanoparticle induced genotoxicity can arise and a distinction between primary and secondary genotoxicity can be made. Inflammation drives secondary toxicity as suggested by} Emma Åkerlund \textit{et al.} \hl{where they showed that conditioned media from differentiated THP1 cells induced DNA-damage in HBEC after} 3h~exposure~\cite{Akerlund2019}. \hl{While there are inflammatory-related pathways in the results for the THP1 cell line, only the high concentration has three DNA-damage related pathways in the results. The presence of DNA-damage related pathways and inflammatory related pathways could indicate secondary genotoxicity for the THP1 cells upon exposure to the high concentration. However, more research is needed to address this exact mechanism.}

@@

\hl{Caco-2/HT29-MTX} cells seem resilient to the exposure and show very little activation of toxicity-related processes. \hl{While for this dataset a co-culture is used,} the passage of nanoparticles through the cellular barriers of \hl{Caco-2} cells has been shown to be limited which could result in reduced uptake and therefore less toxic response~\cite{Ye2017}. \hl{This could partially explain the resilience we see in our results.} Subsequently, this indicates a lower importance of gastro-intestinal uptake in general \hl{but no direct conclusions could be made based on the results and information we have.} On the other hand, TiO\textsubscript{2}-nanobelts caused the biggest toxicological response to THP1 cells,as they are more sensitive to exposure compared to epithelial cell lines RLE-6TN and BEAS-2B~\cite{Xia2013}. Moreover, it has been reported that this response is specific to the nanobelt form of TiO\textsubscript{2}~\cite{Xia2013}.

@@

To illustrate how the effects can be studied in more depth we visualized the log fold change of all six conditions on the Oxidative Stress pathway (wikipathways:WP408~\cite{WP408}), the Apoptosis pathway (wikipathways:WP254~\cite{WP254}) and the DNA Mismatch Repair pathway (wikipathways:WP531~\cite{WP531}). It shows that for most genes in these pathways gene expression data is present (see additional Figures~\ref{fig:figA2} and~\ref{fig:fig3}).

While the Oxidative Stress pathway shown in Figure~\ref{fig:fig3}, only shows up to be significantly overrepresented for the THP1 cell line for the high concentration it is still an interesting pathway to dive deeper into. The NFKB1 gene, which encodes for the Nf$\kappa$B-p105 subunit, has the highest log fold change for the THP1 cell line, high concentration. Additionally, the THP1 cell line, low concentration shows a positive log fold change as well, whereas the SAE cell line shows negative log fold changes and the \hl{Caco-2/HT29-MTX co-culture} noticeably lower positive log fold changes for both concentrations. Furthermore, the SOD2 gene, which is involved in the conversion of superoxide and protects against oxidative stress, shows a similar pattern~\cite{Urso2003}. These genes indicate that the THP1 cell line responds to oxidative stress by increasing the expression of protective genes.

\begin{figure}[ht!]
\includegraphics[height=12cm,keepaspectratio]{fig3.png}
  \caption{Visualization of log fold change on the Oxidative Stress pathway (wikipathways:WP408) all six conditions.
\hl{Cellular models} are depicted from left to right as \hl{Caco-2/HT29-MTX}, SAE and THP1. The top row depicts the low concentrations whereas the bottom row depicts the high concentrations. \hl{Color gradient for differential gene expression after exposure goes from blue (downregulated) via white (not changed) to red (upregulated).}}
\label{fig:fig3}
\end{figure}

Moreover, the Apoptosis pathway shows positive log fold changes of the CASP2 and CASP7 genes, which are involved in apoptosis execution, for the SAE and THP1 cell lines for both concentrations whereas the \hl{Caco-2/HT29-MTX co-culture} shows no negative or positive log fold change~\cite{BouchierHayes2011,Lamkanfi2010}. Furthermore, the pathway shows positive log fold changes for the apoptosis promoting interferon regulatory factors such as IRF4 and IRF5~\cite{Stang2007,Fanzo2003,Hu2009}, which is a similar pattern as described for the CASP2 and CASP7 genes. This could be an indication that apoptosis is stimulated in these cell lines. However, the original study by Tilton \textit{et al.} does not show a significant decrease in cell viability. 
The DNA Mismatch Repair pathway shows a positive log fold change \hl{for the LIG1 gene}, which encodes for DNA ligase 1 \hl{for both the Caco-2/HT29-MTX and THP1 cells}. This enzyme is involved in both DNA replication and in this context more importantly repair~\cite{Mota2019}. The aforementioned genes show positive log fold changes for the THP1 cell line throughout the pathway. The increase in expression of these genes could be an indication that DNA mismatch repair is activated due to exposure to TiO\textsubscript{2}-nanobelts.

\subsection*{The advantage of our approach}
Enrichment analysis for gene expression data is well-established and can be easily automated to increase reproducibility~\cite{Yu2012}. The interpretation of the long list of altered, often overlapping pathways is still a challenge. \hl{To mitigate this challenge we proposed} an automated approach to filter the pathway enrichment result towards a specific biological focus, in this case toxicity. A more context-specific interpretation is facilitated \hl{using GO-term gene sets of interest}. The generated pathway-gene networks showcase the connectedness of the processes and provide a more systemic view than looking at individual pathways. The approach yields a fast and easy method to select pathways of interest for more detailed scrutiny, as we have shown. While interpretation and comparison between multiple datasets on a process level are still challenging, the automated analysis workflow used supports the exploration of the data~\cite{TiO2-scripts}. \hl{Our approach showed the biological response of the three cellular models on a biological process level, making use of pathways. Using pathways to investigate the response does not only highlight biological processes that are affected, but also enables researchers to dive deeper into the pathways on a gene-level. We showed this in our research, where we could easily discuss affected genes in relation to biological processes based on the results we retrieved from our approach. Furthermore, our approach is suitable to quickly retrieve results towards specific biological processes which will aid researchers in their future research. Moreover, our approach is suitable for quick reanalysis each time new pathway knowledge is known, thus pathway databases are updated.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
This study investigated the molecular response of three different cell lines to exposure to TiO\textsubscript{2}-nanobelts. Using our process-level analysis based on pathway analysis \hl{in combination with the use of} gene sets, and network visualization, our findings support the results by S.C. Tilton \textit{et al.} showing that the three \hl{cellular models, Caco-2/HT29-MTX}, SAE, and THP1, show different toxicity-related responses to the exposure of TiO\textsubscript{2}-nanobelts from resilient \hl{Caco-2/HT29-MTX} and SAE cells to strongly responding THP1 cells. The latter is not unexpected since the observed effects align with the biological function of these immune cells. Importantly, the approach allowed us to not \hl{only} find changes in gene expression, but also find responding molecular pathways via the pathway analysis. Additionally, \hl{it} allows us to filter the broad pathway enrichment results to a focused view on the cytotoxic processes affected. \hl{The filtering steps we included in our workflow allow a very targeted approach.} This allows us to visualize and explore the interactions between responding genes, based on underlying molecular processes \hl{in greater detail and in a less time-consuming manner.  The approach is suitable for quick reanalysis of datasets each time pathways databases are updated with new found pathway knowledge. Moreover,} This versatile approach captured in the R-script can be easily adapted to isolate other processes by using other Gene Ontology terms diverting the focus to other biological processes of interest.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}

@@

The dataset consisted of 18 samples from three human-like \hl{cellular models}, \textit{i.e.} \hl{Caco-2/HT29-MTX}, SAE, and THP1, which were exposed to either medium (control), 10~$\mu$g/ml or 100~$\mu$g/ml TiO\textsubscript{2}-nanobelts for 24 hours in triplicate. \hl{To run the same analysis for the 1 hour time point the workflow in the 1h repository can be} used~\cite{TiO2-scripts-1h}. \hl{The number of samples, cellular models and number of replicates were the same for the 1 hour time point compared to the 24 hours time point.} Culture conditions were kept as identical as possible between the three cell lines \hl{for both time points}. Details can be found via the original publication by Tilton~\textit{et al.}~\cite{Tilton2013}.

@@

\subsection*{Toxicity-related processes}
Within the two human pathway collections used, toxicity-related pathways were identified based on the gene overlap with the toxicity-related gene sets. The overlap was calculated by dividing the number of toxicity-related genes present in the pathway by total number of genes present in the pathway. The gene sets were retrieved from the Gene Ontology (GO, version: release 2020-06) for the following four toxicity-related GO-terms: “apoptotic process” (GO:0006915), “inflammatory response” (GO:0006954), “cellular response to DNA damage stimulus” (GO:0006974) and “response to oxidative stress” (GO:0006979)~\cite{Ashburner2000,Carbon2020}. Associated genes were retrieved using the biomaRt package in R (version 2.42.0, Ensembl Genes 100)~\cite{Durinck2005,Durinck2009}. Using the GO evidence codes, only genes with experimental evidence or manually curated annotations were included to ensure high confidence that the gene is involved in the specific process (IBA, IC, IDA, IEP, IGI, IMP, IPI, TAS, http://geneontology.org/docs/guide-go-evidence-codes/, see additional Table~\ref{tab:tabA2}). Gene overlap between GO-terms was visualized using Venny version 2.1.0~\cite{Venny}.

@@

\noindent 
\begin{tabular}{@{}ll}
TiO\textsubscript{2} & titanium dioxide\\
SAE & small airway epithelial cells\\
THP1 & human monocytic cells\\
\hl{Caco-2} & human epithelial colorectal adenocarcinoma cells\\
GO & Gene Ontology\\
\end{tabular}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Optional
\newpage
%\appendixtitles{no} % Leave argument "no" if all appendix headings stay EMPTY (then no dot is printed after "Appendix A"). If the appendix sections contain a heading then change the argument to "yes".
\appendixstart
\appendix
\section{}
